---
title: "Pig data in Uganda for Michel"
author: "Catherine and Stephen"
date: "23 October 2017"
output: html_document
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE, cache=T)
```


# The data

We are extracting the DHS standard survey of 2011. We have extracted
1. toilet that a household use
2. number of livestock owned by the households
3. a couple of household charateristics

```{r dataprep, echo = F, warning = F, message = F}
path<-'D:/Dropbox/Dione'
setwd(path)
library(foreign)
library(maptools)
library(ggplot2)
library(gstat)
library(raster)
library(colorspace)
library(rgeos)
data<-read.spss('UGA_2012_60/HR.SAV', to.data.frame = T,use.value.labels = FALSE)
geo<-readShapePoints('UGA_2012_60/GE.shp')
shp<-readShapePoly('UGA_2012_60/sdr_subnational_boundaries.shp')
admin<-readShapePoly('admin-gdam/UGA_adm1.shp')
admin$NAME_1<-as.character(admin$NAME_1)
admin$label<-NA
admin$label[(admin$NAME_1=='Lira'|admin$NAME_1=='Masaka'|admin$NAME_1=='Mukono'|admin$NAME_1=='Kamuli'|admin$NAME_1=='Hoima'|admin$NAME_1=='Moyo'|admin$NAME_1=='Kumi'|admin$NAME_1=='Apac'|admin$NAME_1=='Kaberamaido'|admin$NAME_1=='Soroti'|admin$NAME_1=='Kayunga')]<-admin$NAME_1[(admin$NAME_1=='Lira'|admin$NAME_1=='Masaka'|admin$NAME_1=='Mukono'|admin$NAME_1=='Kamuli'|admin$NAME_1=='Hoima'|admin$NAME_1=='Moyo'|admin$NAME_1=='Kumi'|admin$NAME_1=='Apac'|admin$NAME_1=='Kaberamaido'|admin$NAME_1=='Soroti'|admin$NAME_1=='Kayunga')]
shp2<-fortify(shp)
admin2<-fortify(admin)
#data$HV246G <-as.numeric(data$HV246G )
data$Pg_num<-ifelse(data$HV246G >95.5,NA,data$HV246G)
data$Pg_h<-ifelse (data$Pg_num>0,1,0)
data$Pg_2<-ifelse (data$Pg_num==1|data$Pg_num==2&data$HV025=='Rural',1,0)
data$wc1<-ifelse(data$HV205=='31'|data$HV205=='25'|data$HV205=='24',1,0)
data$wc2<-ifelse(data$HV205=='31',1,0)
data$w1<-ifelse(data$HV270=='1',1,0)
data$w2<-ifelse(data$HV270=='2',1,0)
data$w3<-ifelse(data$HV270=='3',1,0)
data$w4<-ifelse(data$HV270=='4',1,0)
data$w5<-ifelse(data$HV270=='5',1,0)

myClu = geo@data
myClu<-myClu[which(!myClu$LONGNUM==0),]

myClu = myClu[,c("DHSCC","DHSCLUST","DHSREGCO","DHSREGNA","URBAN_RURA", "ADM1NAME", "LONGNUM","LATNUM")]
names(myClu)=c("Cc","HV001","REGCODE", "HV024","IsRural", "AdName","x","y")



temp<- aggregate(x=data$Pg_num, by=list(data$HV001), 'mean', na.rm = T)
myPosVec = match(myClu$HV001,temp$Group.1 )
myClu$Pg_num = temp$x[myPosVec]

temp<- aggregate(x=ifelse(data$Pg_num==0,NA,data$Pg_num), by=list(data$HV001), 'mean', na.rm = T)
myPosVec = match(myClu$HV001,temp$Group.1 )
myClu$Pg_hs = temp$x[myPosVec]

temp<- aggregate(x=data$Pg_h, by=list(data$HV001), 'mean', na.rm = T)
myPosVec = match(myClu$HV001,temp$Group.1 )
myClu$Pg_h = temp$x[myPosVec]  

temp<- aggregate(x=data$Pg_2, by=list(data$HV001), 'mean', na.rm = T)
myPosVec = match(myClu$HV001,temp$Group.1 )
myClu$Pg_2 = temp$x[myPosVec]    

temp<- aggregate(x=data$wc1, by=list(data$HV001), 'mean', na.rm = T)
myPosVec = match(myClu$HV001,temp$Group.1 )
myClu$wc1 = temp$x[myPosVec]  


temp<- aggregate(x=data$wc2, by=list(data$HV001), 'mean', na.rm = T)
myPosVec = match(myClu$HV001,temp$Group.1 )
myClu$wc2 = temp$x[myPosVec]   


temp<- aggregate(x=data$w1, by=list(data$HV001), 'mean', na.rm = T)
myPosVec = match(myClu$HV001,temp$Group.1 )
myClu$w1 = temp$x[myPosVec]   


temp<- aggregate(x=data$w2, by=list(data$HV001), 'mean', na.rm = T)
myPosVec = match(myClu$HV001,temp$Group.1 )
myClu$w2 = temp$x[myPosVec]   

temp<- aggregate(x=data$w3, by=list(data$HV001), 'mean', na.rm = T)
myPosVec = match(myClu$HV001,temp$Group.1 )
myClu$w3 = temp$x[myPosVec]   

temp<- aggregate(x=data$w4, by=list(data$HV001), 'mean', na.rm = T)
myPosVec = match(myClu$HV001,temp$Group.1 )
myClu$w4 = temp$x[myPosVec]   

temp<- aggregate(x=data$w5, by=list(data$HV001), 'mean', na.rm = T)
myPosVec = match(myClu$HV001,temp$Group.1 )
myClu$w5 = temp$x[myPosVec]   
```

## Livestock ownership

This map looks at housholds that have reported pigs
```{r pressure, echo=FALSE}
ggplot() + geom_polygon(data=admin2, aes(x=long, y=lat, group=group), fill="grey80", colour="grey90", alpha=1) + geom_point(data=myClu, aes(x=x, y=y,color=Pg_h) ) + scale_colour_gradientn(colours=heat.colors(12))+    coord_equal(ratio=1)
```


```{r kpressure, echo=FALSE}
grid<-makegrid(shp, cellsize=0.0083)
coordinates(grid)<-~x1+x2
geodata<-myClu
coordinates(geodata)<- ~x+y
vgm = variogram(Pg_h~1, geodata)
fit = fit.variogram(vgm, model = vgm(0.03, "Sph", 2, 0.01))
plot(vgm, fit)
kriged = krige(Pg_h~1, geodata, grid, model = fit)
ggplot(data=kriged@data, aes(x=kriged$x1, y=kriged$x2)) + geom_tile(aes(fill=kriged$var1.pred)) + coord_equal() +
  scale_fill_gradient(low = "yellow", high="red") +
  geom_polygon(data=admin2, aes(x=long, y=lat, group=group), color="black", alpha=0)+ theme_bw()

```

This sections looks at flock sizes from households who have reported owning pigs 
```{r hs, echo=FALSE}
ggplot() + geom_polygon(data=admin2, aes(x=long, y=lat, group=group), fill="grey80", colour="grey90", alpha=1) + geom_point(data=myClu, aes(x=x, y=y,color=Pg_hs) ) + scale_colour_gradientn(colours=heat.colors(12))+    coord_equal(ratio=1)
```

```{r kpressure2, echo=FALSE}
idw = idw(Pg_hs ~1, geodata[!is.na(geodata$Pg_hs),],grid, maxdist = Inf, idp = 2)
ggplot(data=idw@data, aes(x=idw$x1, y=idw$x2)) + geom_tile(aes(fill=idw$var1.pred)) + coord_equal() +
  scale_fill_gradient(low = "yellow", high="red") +
  geom_polygon(data=admin2, aes(x=long, y=lat, group=group), color="black", alpha=0)+ theme_bw()
```

in order to compute pig density we do the following calculation : 
population (from world pop 2015 unadjusted) x percent of household keeping pigs x flock size

```{r pop, echo=FALSE}
pop<-raster('pop/UGA_ppp_v2b_2015.tif')

gridded(kriged)<-T
Pg_h_r <- raster(kriged, "var1.pred")
Pg_h_r<-mask(Pg_h_r,shp)
gridded(idw)<-T
Pg_hs_r<-raster(idw, "var1.pred")
Pg_hs_r<-mask(Pg_hs_r,shp)
pop<-resample(pop,Pg_h_r)
Pg_den<-pop*10*Pg_h_r*Pg_hs_r#*10 because of the resolution change
plot(Pg_den)
Pg_den_co<-Pg_den
Pg_den_co[Pg_den_co>100]<-NA
plot(Pg_den_co)
```

This map does not show much variation because there is an overestimation of pig density in urban areas. So we re-run it showing taking the extreme value out, i.e values above 50. Then a pattern appears.
Let us compare it with the FAO Robinson pig distribution map. (my laptop does not manage to do this,i'll rerun from office)


```{r  distpop, echo=F}
#this is robinson 2006
# ppop<-raster('H:/GIS data/global/livestock_data/PIGS/Glb_Pigs_CC2006_AD.tif')
# ppop<-mask(ppop,shp)
# ppop<-crop(ppop,shp)
# ppop<-resample(ppop,Pg_h_r)
# writeRaster(ppop,'ppop.tif', overwrite=T)
ppop<-raster('ppop.tif')
breakpoints <- c(0, 1, 2, 5, 10, 25, 75, 250,1000, 3000)
colors <- c("gray90",rainbow_hcl(7))
plot(ppop, breaks = breakpoints, col= colors)
plot(admin, add=T)
```



## extensive system 
This these are places where people let there pigs roam and therefore there is high potential for diseases 

```{r extensive, echo=FALSE}
ggplot() + geom_polygon(data=admin2, aes(x=long, y=lat, group=group), fill="grey80", colour="grey90", alpha=1) + geom_point(data=myClu, aes(x=x, y=y,color=Pg_2) ) + scale_colour_gradientn(colours=heat.colors(12))+    coord_equal(ratio=1)
```

```{r extensivek, echo=FALSE}
vgm = variogram(Pg_2~1, geodata)
fit = fit.variogram(vgm, model = vgm(0.03, "Sph", 2, 0.01))
plot(vgm, fit)
kriged = krige(Pg_2~1, geodata, grid, model = fit)
ggplot(data=kriged@data, aes(x=kriged$x1, y=kriged$x2)) + geom_tile(aes(fill=kriged$var1.pred)) + coord_equal() +
  scale_fill_gradient(low = "yellow", high="red") +
  geom_polygon(data=admin2, aes(x=long, y=lat, group=group), color="black", alpha=0)+ theme_bw()
gridded(kriged)<-T
Pg_2<-raster(kriged,"var1.pred")
```


## sanitation
These maps show the percentage of household with low sanitation facilities
The first map is about household who have no facilities at all 
```{r sanitation, echo=FALSE}
ggplot() + geom_polygon(data=admin2, aes(x=long, y=lat, group=group), fill="grey80", colour="grey90", alpha=1) + geom_point(data=myClu, aes(x=x, y=y,color=wc2) ) + scale_colour_gradientn(colours=heat.colors(12))+    coord_equal(ratio=1)
```


```{r sank1, echo=FALSE}
vgm = variogram(wc2~1, geodata)
fit = fit.variogram(vgm, model = vgm(0.03, "Sph", 2, 0.01))
plot(vgm, fit)
kriged = krige(wc2~1, geodata, grid, model = fit)
ggplot(data=kriged@data, aes(x=kriged$x1, y=kriged$x2)) + geom_tile(aes(fill=kriged$var1.pred)) + coord_equal() +
  scale_fill_gradient(low = "yellow", high="red") +
  geom_polygon(data=admin2, aes(x=long, y=lat, group=group), color="black", alpha=0)+ theme_bw()
gridded(kriged)<-T
wc2<-raster(kriged,"var1.pred")
```



And in this one with no sanitation or uncovered facilities
```{r sanitation4, echo=FALSE}
ggplot() + geom_polygon(data=admin2, aes(x=long, y=lat, group=group), fill="grey80", colour="grey90", alpha=1) + geom_point(data=myClu, aes(x=x, y=y,color=wc1) ) + scale_colour_gradientn(colours=heat.colors(12))+    coord_equal(ratio=1)
```

```{r sank2, echo=FALSE}
vgm = variogram(wc1~1, geodata)
fit = fit.variogram(vgm, model = vgm(0.03, "Sph", 2, 0.01))
plot(vgm, fit)
kriged = krige(wc1~1, geodata, grid, model = fit)
ggplot(data=kriged@data, aes(x=kriged$x1, y=kriged$x2)) + geom_tile(aes(fill=kriged$var1.pred)) + coord_equal() +
  scale_fill_gradient(low = "yellow", high="red") +
  geom_polygon(data=admin2, aes(x=long, y=lat, group=group), color="black", alpha=0)+ theme_bw()
gridded(kriged)<-T
wc1<-raster(kriged,"var1.pred")
```

## poverty level
based on percentage of household in the poorest 20%

```{r sanitation3, echo=FALSE}
ggplot() + geom_polygon(data=admin2, aes(x=long, y=lat, group=group), fill="grey80", colour="grey90", alpha=1) + geom_point(data=myClu, aes(x=x, y=y,color=w1) ) + scale_colour_gradientn(colours=heat.colors(12))+    coord_equal(ratio=1)
```

```{r wealth1, echo=FALSE}
vgm = variogram(w1~1, geodata)
fit = fit.variogram(vgm, model = vgm(0.03, "Wav", 2, 0.01))
plot(vgm, fit)
kriged = krige(w1~1, geodata, grid, model = fit)
ggplot(data=kriged@data, aes(x=kriged$x1, y=kriged$x2)) + geom_tile(aes(fill=kriged$var1.pred)) + coord_equal() +
  scale_fill_gradient(low = "yellow", high="red") +
  geom_polygon(data=admin2, aes(x=long, y=lat, group=group), color="black", alpha=0)+ theme_bw()
gridded(kriged)<-T
w1<-raster(kriged,"var1.pred")
```
percentage of poor in the 40% percent poorest category (poorest and poor)
```{r wealth2a, echo=FALSE}
ggplot() + geom_polygon(data=admin2, aes(x=long, y=lat, group=group), fill="grey80", colour="grey90", alpha=1) + geom_point(data=myClu, aes(x=x, y=y,color=(w1+w2) )) + scale_colour_gradientn(colours=heat.colors(12))+    coord_equal(ratio=1)
```

```{r wealth2, echo=FALSE}
vgm = variogram((w1=w2)~1, geodata)
fit = fit.variogram(vgm, model = vgm(0.03, "Sph", 2, 0.01))
plot(vgm, fit)
kriged = krige((w1+w2)~1, geodata, grid, model = fit)
ggplot(data=kriged@data, aes(x=kriged$x1, y=kriged$x2)) + geom_tile(aes(fill=kriged$var1.pred)) + coord_equal() +
  scale_fill_gradient(low = "yellow", high="red") +
  geom_polygon(data=admin2, aes(x=long, y=lat, group=group), color="black", alpha=0)+ theme_bw()
gridded(kriged)<-T
w2<-raster(kriged,"var1.pred")
```

# Investigating the distribution of the key variables to define meaningful breaks into high and low
## livestock ownership distribution
for the final map we are using the 3rd quantile, i.e. the map shows were the 25% clusters with most households with pigs are. 
```{r distr, echo=T}
hist(Pg_h_r@data@values)
summary(Pg_h_r@data@values)
Pg_h_b<-Pg_h_r
Pg_h_b[Pg_h_r>0.23]<-1
Pg_h_b[Pg_h_b<0.23]<-0
plot(Pg_h_b)
```
## pig population 
based on DHS computation
```{r distr2, echo=T}
hist(Pg_den@data@values)
summary(Pg_den@data@values)
Pg_den_b<-Pg_den
Pg_den_b[Pg_den>1]<-1
Pg_den_b[Pg_den_b<1]<-0
plot(Pg_den_b)
plot(admin,add=T)
```
based on FAO Robinson layer (this does not work yet)
```{r distr3, echo=T}
hist(ppop@data@values)
summary(ppop@data@values)
Pg_den_b2<-ppop
Pg_den_b2[ppop>1]<-1
Pg_den_b2[Pg_den_b2<1]<-0
plot(Pg_den_b2)
plot(admin,add=T)
```

## pig extensive rural system 
```{r distr4, echo=T}
hist(Pg_2@data@values)
summary(Pg_2@data@values)
Pg_2_b<-Pg_2
Pg_2_b[Pg_2>0.153]<-1
Pg_2_b[Pg_2<0.153]<-0
plot(Pg_den_b)
```
## poverty the 40% poorest

```{r distr5, echo=T}
hist(w2@data@values)
summary(w2@data@values)
w2_b<-w2
w2_b[w2>0.666]<-1
w2_b[w2_b<0.666]<-0
plot(w2_b)
```


## uncovered sanitation

```{r distr6, echo=T}
hist(wc1@data@values)
summary(wc1@data@values)
wc1_b<-wc1
wc1_b[wc1>0.408]<-1
wc1_b[wc1_b<0.408]<-0
plot(wc1_b)
```


## poverty the 40% poorest

```{r distr7, echo=T}
hist(w1@data@values)
summary(w1@data@values)
w1_b<-w1
w1_b[w1>0.438]<-1
w1_b[w1_b<0.438]<-0
plot(w1_b)
plot(admin,add=T)
```

#First overlays 
## proportion of households with bad sanitation, proportion of extensive households and poverty 

* A proportion of households bad sanitation high =1 is low = 0
* B proportion of poor 40% is high = 1.01 is low zero 
* C proportion of extensive households is high =1.1 is low =0 

1. 0 = ABC low 
2. 1 = A high (yellow)
3. 1.01 = B high (blue)
4. 1.1 = C high (red)
5. 2.01= AB high (green)
6. 2.1 = AC high (orange)
7. 2.11= BC high (purple)
8. 3.11= ABC high (brown)

```{r overlay1, echo=T}
breakpoints <- c(0, 0.99, 1.0099, 1.099, 2.0099, 2.099, 2.1099, 3.1099,3.12)
breakpoints2 <- c('all low','A' , 'B', 'C', 'AB', 'AC', 'BC', 'ABC')
colors <- c( "gray90","gold","steelblue1","red","springgreen",'orange','plum1', "brown")

over1<-wc1_b+Pg_2_b*1.1+w2_b*1.01
over1<-mask(over1,admin)
plot(over1,breaks=breakpoints,col=colors, legend=FALSE)
legend("bottomleft", inset=.02, title="classes", legend = breakpoints2
   , fill  =colors, horiz=F, cex=0.8)
plot(admin,add=T)

```

 

## proportion of household with bad sanitaiton, proportion of extensive households and pig distribution (Robinson) 

* A proportion of households bad sanitation high =1 is low = 0
* B pig density   is high  = 1.01 is low zero (robinson)
* C proportion of extensive households is high =1.1 is low =0 


```{r overlay2, echo=T}
over2<-wc1_b+Pg_2_b*1.1+Pg_den_b2*1.01
plot(over2,breaks=breakpoints,col=colors, legend=FALSE) 
legend("bottomleft", inset=.02, title="classes", legend = breakpoints2
   , fill=colors, horiz=F, cex=0.8)
plot(admin,add=T)
```



## proportion of household with bad sanitaiton, proportion of extensive households and pig distribution (Robinson) 

* A proportion of households bad sanitation high =1 is low = 0
* B pig density   is high  = 1.01 is low zero 
* C proportion of extensive households is high =1.1 is low =0 


```{r overlay3, echo=T}
over3<-wc1_b+Pg_2_b*1.1+w1_b*1.01
plot(over3,breaks=breakpoints,col=colors, legend=FALSE) 
legend("bottomleft", inset=.02, title="classes", legend = breakpoints2
   , fill=colors, horiz=F, cex=0.8)
plot(admin,add=T)

```


* A proportion of households bad sanitation high =1 is low = 0
* B pig density   is high  = 1.01 is low zero  from DHS
* C proportion of extensive households is high =1.1 is low =0 


```{r overlay4, echo=T}
over4<-wc1_b+Pg_2_b*1.1+Pg_den_b*1.01
plot(over4,breaks=breakpoints,col=colors, legend=FALSE) 
legend("bottomleft", inset=.02, title="classes", legend = breakpoints2
   , fill=colors, horiz=F, cex=0.8)
plot(admin,add=T)

```

* A proportion of households bad sanitation high =1 is low = 0
* B pig density   is high  = 1.01 is low zero robinson
* C proportion of poor 40% is high = 1.1 is low zero 

This is the final one 
```{r overlay5, echo=T}
breakpoints <- c(0, 0.99, 1.0099, 1.099, 2.0099, 2.099, 2.1099, 3.1099,3.12)
breakpoints2 <- c('all low','A' , 'B', 'C', 'AB', 'AC', 'BC', 'ABC')
colors <- c( "gray90","gold","steelblue1","red","springgreen",'orange','plum1', "brown")

over5<-wc1_b+Pg_den_b2*1.01+w2_b*1.1
plot(over5,breaks=breakpoints,col=colors, legend=FALSE) 
legend("bottomleft", inset=.02, title="classes", legend = breakpoints2
   , fill=colors, horiz=F, cex=0.8)
plot(admin,add=T)
invisible(text(getSpPPolygonsLabptSlots(admin), labels=as.character(admin$label), cex=0.6))

writeRaster(wc1_b,'wc1_b.tif')
writeRaster(Pg_den_b2,'Pg_den_b2.tif')
writeRaster(w2_b,'w2_b.tif')

writeRaster(wc1,'wc1.tif')
writeRaster(Pg_den,'Pg_den.tif')
writeRaster(w2_b,'w2.tif')


```







